
# Use R efficiently with data.table!
<!-- for very fast and user friendly data processing -->


 ## TL;DR
 
 1. For any code you do (ie, by default) your first lines of code should be:
 ```{r}
 
 
 ```
 

The `data.table` package developed by Matt Dowle is a game changer for many data scientists

Learn about it...  and use it always, 
by default:
```{r}
library(data.table)
dtIris <- data.table(iris) # or
df <- iris; dtIris <- setDT(df)
```


Where to learn: 

- https://github.com/Rdatatable/data.table

- https://github.com/chuvanan/rdatatable-cookbook

 

- http://r-datatable.com (https://rdatatable.gitlab.io/data.table/)
- https://www.datacamp.com/courses/time-series-with-datatable-in-r
- https://www.datacamp.com/courses/data-manipulation-in-r-with-datatable
- https://github.com/Rdatatable/data.table/wiki/Articles
- https://rpubs.com/josemz/SDbf - Making .SD your best friend
 



## data.table vs. dplyr


data.table (Computer language) way  vs. dplyr ("English language") way  

A) The best: No wasted computations .No new memory allocations.
dtLocations %>% .[WLOC == 4313, WLOC:=4312]
B) No new memory allocations, but computations are done with ALL rows.
dtLocations %>% .[, WLOC:=ifelse(WLOC==4313, 4312, WLOC)]
A) The worst: Computations are done with ALL rows. Furthermore, the entire data is copied from  one memory location to another.  (Imagine if your data as in 1 million of cells, of which only 10 needs to be changed !)
dtLocations <- dtLocations %>%   mutate(WLOC=ifelse(WLOC==4313, 4312, WLOC))
NB:  dtLocations %>%  . [] is the same as dtLocations[]. so you can use it in pipes.

 

<!-- Conclusion: Use data.table for speed and efficient coding instead of dplyr (i.e.tibbles)! -->


<!-- Comment  -->

<!-- je côtoie de nombreux scientifiques en environnement qui sont nouveaux dans le monde de la programmation et qui n'ont pas un gros volume de données. Pour eux, dplyr est plus « accessible » que data.table. -->

<!-- Un jour, ils y viendront peut-être, mais en attendant, je vais continuer à enseigner les rudiments de R en proposant dplyr et en leur parlant brièvement du package data.table. -->


## Extensions of data.table

There's considerable effort to marry  data.table package with dplyr package. Here are  notable ones:

- https://github.com/tidyverse/dtplyr (Version: 1.1.0, Published: 2021-02-20, From Hadley himself - I found it quite cumbersome still though...)
- https://github.com/asardaes/table.express (Version: 0.3.1 Published: 2019-09-07 - somewhat easier?)
- https://github.com/markfairbanks/tidytable (Version: 0.6.2 Published: 2021-05-18 - seems to be the best supported of the three?)





# Python and R unite!

Check out our latest Lunch and Learn Meetup  : 
'Lunch and Learn' Meetup: Dual Coding - Python and R unite! : GCcollab

 

It shows how to Execute R from Python and Vice Versa, inspired by this blog: this https://www.kdnuggets.com/2015/10/integrating-python-r-executing-part2.html

 

The used code is available from GCCode:  https://gccode.ssc-spc.gc.ca/r4gc/resources/howto




# From Excel to R

There was a keen interest expressed at last Friday meetup on transitioning from Excel to R. Incidentally, there was an RStudio Community Meet-up focused exactly on this topic:  Meetup: Making the Shift from Excel to R: Perspectives from the back-office

<!-- ## 1 -->

Many Q&A from this meetup are posted here:

https://community.rstudio.com/t/meetup-making-the-shift-from-excel-to-r-perspectives-from-the-back-office/100467

Below are listed some. There's also a link to a video-recording of it! (at link above)


```
What was the most common problem when moving from Excel to R? [Excel -> R Meetup Q&A]
R Markdown vs R Scripts in Enterprise Workflows [Excel -> R Meetup Q&A]
What sort of packages did you use for reconciliation? [Excel -> R Meetup Q&A]
What is the best way to share the results of an analysis that you’ve made in R? [Excel -> R Meetup Q&A]
How do you get buy-in transitioning from Excel to R? [Excel -> R Meetup Q&A]
How do you get buy-in from the IT team around R package security concerns? [Excel -> R Meetup Q&A]
How do you move data from tricky worksheets into R? [Excel -> R Meetup Q&A]
With so many ways of doing the same things, how do you find the best way to do something? [Excel -> R Meetup Q&A]
```


<!-- ## 2 -->


This dictionary of Excel/R equivalent was very useful is finding a starting point for common functions:     
https://paulvanderlaken.com/2018/07/31/transitioning-from-excel-to-r-dictionary-of-common-functions/  

 

For people who are looking for something more comprehensive, this website is very useful:   
https://rstudio-conf-2020.github.io/r-for-excel/



<!-- ## 3 -->

Pour ma part, j'adhère à l'approche proposée par The [Carpentries](https://datacarpentry.org/spreadsheet-ecology-lesson/00-intro/index.html), soit l'utilisation d'Excel pour la saisie de données et une partie du contrôle de la qualité. Par la suite, les données sont exportées en csv pour être importées dans R.




# Other tips for  Efficient coding in R 
<!-- (coding style, memory-efficient coding, collaboration-ready codes, source control) -->

<!-- TBA -->

<!-- ## Debugging tricks -->

<!-- TBA -->

## Variable names !

Variable names  need to help you every time you see them! 

Examples:
```{r}


```



### Code formating

Install 'slyler' package to have RStudio addin to automate formatting your code




## Code starter tricks


### Most important / useful libraries

These are libraries / codes that one could use by default when starting  a new code.

```{r}

# 0. General libraries and functions ----

# options(digits = 3); # 7
# options(max.print = 100) # 1000
# options(scipen=999); #remove scientific notation

library(magrittr)
# library(readxl)
# library(readr)
# library(stringr)

## library(tidyverse) # includes:  readr ggplot2 stringr (which we use), dplyr tidyr tibble forcats purrr (may use later) and others - https://www.tidyverse.org/packages/

# library(dplyr) # or
#library(dtplyr)

# library(ggplot2) 
# library(lubridate,  quietly=T) # to work work times and dates
# options(lubridate.week.start =  1)

library(data.table) # needs to be AFTER lubridate
options(datatable.print.class=TRUE)

# library(dygraphs)
# library(plotly)
# library(DT)

`%ni%` = Negate(`%in%`)
```

## Efficient workflows


<!-- Commonly, there are two possible  -->

### Workflow: Data-first approach

<!-- ### Starting with new Data set -->


So, you found a new
<!-- previously not seen  -->
Open Data
dataset. Great - now you can do something with it!  This is how you  start .

```{r}

```




### Workflow: Task/needs/algorithm-first approach

TBA


## Object oriented programming in R



https://techvidvan.com/tutorials/r-object-oriented-programming/


### S3


### R6 







## RStudio tricks



### Coding online

Platforms for coding R online   
<!-- a) https://rextester.com/l/r_online_compiler -->
<!-- b) https://repl.it/languages/rlang (no login required) -->
 https://rstudio.cloud/ (login required) - All in One place for learning and coding



#### Running multiple RStudio versions

You want to be able to run  multiple versions of RStudio in Windows? You can do with the following executable .bat script.

```
# Run-RStudio-1.4.bat

@echo off
title Starting RStudio!
echo Hello Dear,
echo Starting RStudio 1.4 (from C:\Users\gxd006\Downloads\RStudio-1.4.1106\bin\) for you...

C:\Users\abc123\Downloads\RStudio-1.4.1106\bin\rstudio.exe

Set WshShell = CreateObject("WScript.Shell")
WshShell.Run chr(34) & "C:\Users\gxd006\Downloads" & chr(34), 0
Set WshShell = Nothing

```






# Using R with GC data infastructure (gcdocs, AWS, etc)


## gcdocs

Q:

I'm wondering if anyone has had any success accessing data from GCDocs in your respective departments ? I believe that GCDocs is implemented across most (if not all) departments so wondering if there are any existing solutions to read/write data from it.



Also wondering about whether any of you have had any luck accessing Microsoft 365 via R as well? I've had success with Microsoft365R package (https://github.com/Azure/Microsoft365R) from a personal point of view but it doesn't play well with business accounts.


<!-- A: -->
<!--   I tried using the Microsoft365R package to access my departmental (ECCC) email without success. When I tried to access it, a window popped up allowing me to request access authorization so I clicked the "Submit" (or whatever) button. That was weeks ago. Never heard anything more about it. -->

<!-- After trying many ways and weeks, I found this is NOT possible. -  It runs internal code on gcdoc end  that validates that you have right to access it and then, if you do, it also logs your action within the document's "Audit" attribute.  -->
<!-- So we still have to always make a local copy of the data (manually!), and only then we can process it from R. -->


